% SPDX-FileCopyrightText: Copyright (c) 2025-2026 Objectionary.com
% SPDX-License-Identifier: MIT

The Stream API\footnote{\url{https://openjdk.org/projects/jdk8/}} introduced in Java~8 makes stream processing more expressive~\citep{kiselyov2017}.
However, it relies heavily on virtual calls to temporary objects.
Consequently, streams may work slower than procedural imperative loops~\citep{khatchadourian2020study}.

Stream fusion may help by reducing stream operations and virtual calls~\citep{hirzel2014catalog}.
Consider the following Java stream pipeline:
\begin{ffcode}
var NN = new int[] {1, 1, 2, 3, 5, 8, 13};
var s = IntStream.of(NN) (*@\label{ln:stream-start}@*)
  .map(x -> (*@\highlight{x * x}@*))
  .map(x -> (*@\highlight{x + 1}@*))
  .sum(); (*@\label{ln:stream-end}@*)
\end{ffcode}
It can be optimized by joining two \ff{map} operations into one:
\begin{ffcode}
var s = IntStream.of(NN).map(
  x -> (*@\highlight{x * x + 1}@*)).sum();
\end{ffcode}
It can be further unrolled into an imperative loop:
\begin{ffcode}
var s = 0;
for (var i = 0; i < NN.length; i++)
  s += NN[i] * NN[i] + 1;
\end{ffcode}

The first ahead-of-time (AOT) bytecode-to-bytecode optimizer called \streamliner{} was introduced by \citet{moller2020eliminating}.
It \emph{unrolls} some stream pipelines by transforming them into loops, thus eliminating virtual calls.
However, this approach has limitations.
For example, objects may not escape from lambda expressions.
Additionally, streams may not be assigned to variables or passed as method arguments.
Moreover, \streamliner{} increased performance in older JVMs, but we couldn't reproduce similar results in JVMs newer than 17.
Our experiments showed that \streamliner{} may \emph{reduce} performance of short pipelines in recent JVM versions.

In Java~16, the Stream API was extended with the \ff{mapMulti} method, which works similarly to \ff{flatMap} but is an order of magnitude faster.
We implemented stream \emph{fusion} by combining pairs of \ff{map} and \ff{filter} operations into a single \ff{mapMulti} operation.
We don't replace streams with loops but merge multiple stream operations into a single one.
The snippet at the lines \ref{ln:stream-start}--\ref{ln:stream-end} would look like this, after our optimization:

\begin{ffcode}
var s = IntStream.of(NN).mapMulti(
  (x, c) -> { (*@\highlight{c.accept(x * x + 1)}@*); }
).sum();
\end{ffcode}

Instead of direct bytecode manipulation, our optimization pipeline is implemented as a rewriting system.
We disassemble bytecode to \phic{} expressions~\citep{bugayenko2021eolang}, rewrite them, and assemble into bytecode.
The usage of \phic{} enabled bytecode manipulation through declarative pattern matching and rewriting.

We released \hone{} to Maven Central, an open source Maven plugin that implements our method%
  \footnote{\anon[URL anonymized]{\url{https://github.com/objectionary/hone-maven-plugin}}}.
Any Java project may optimize its bytecode after compilation by adding our plugin to their Maven or Gradle build configuration.
The plugin requires no modifications to source code.

We asked the following research questions:
\requ{1}: Can replacing sequences of \ff{map} and \ff{filter} operations with a single \ff{mapMulti} operation provide performance gains similar to stream unrolling?
\requ{2}: Can stream fusion via \ff{mapMulti} be applied safely, not disrupting the functionality of existing stream pipelines?

Our results were as follows:
On two of nine benchmarks used by \citet{moller2020eliminating}, our optimization outperformed \streamliner{}.
On five benchmarks, our optimization matched \streamliner{}.
Our method didn't optimize two benchmarks because they use terminal stream operations that our method is not intended to optimize.
Optimization of Apache Kafka bytecode didn't cause any of its 31,799 unit tests in 1,511 test suites to fail.

Based on the results, we can provide the following answers to our RQs:
First, stream fusion via \ff{mapMulti} can provide performance gains comparable to or better than stream unrolling.
Second, our optimization has fewer applicability limitations in production Java systems compared with \streamliner{}.

In summary, the contributions of this paper are:
\begin{inparaenum}[1)]
  \item We propose a formal intermediate representation for Java bytecode based on \phic{}, enabling systematic stream fusion via rewrite rules and bridging low-level bytecode with high-level algebraic optimization;
  \item We present \jeo{} v\input{_env/jeo-version.tex}, a Maven plugin for bidirectional mapping between Java bytecode and \phic{} expressions%
      \footnote{\anon[URL anonymized]{\url{https://github.com/objectionary/jeo-maven-plugin}}};
    a rewriting system with 44 rules that replaces \ff{map} and \ff{filter} operations with \ff{mapMulti} operations;
    \phino{} v\input{_env/phino-version.tex}, a tool that applies rewriting rules to \phic{} expressions%
      \footnote{\anon[URL anonymized]{\url{https://github.com/objectionary/phino}}};
    and \hone{} v\input{_env/hone-version.tex}, a Maven plugin that automatically optimizes bytecode \ff{.class} files after compilation;
  \item We report an experimental evaluation showing that our method of stream fusion
    may be more effective on some benchmarks than \streamliner{}, especially on modern JVMs.
  \item We demonstrate by experiment that our method doesn't disrupt
    the functionality of 31,799 unit tests in Apache Kafka.
\end{inparaenum}
