% SPDX-FileCopyrightText: Copyright (c) 2025-2026 Objectionary.com
% SPDX-License-Identifier: MIT

Our method requires JDK~16 or newer due to \ff{mapMulti}, is hardwired to the Stream API tested on JDK~24, and currently transforms only \ff{map} and \ff{filter} among the eleven intermediate operations (\ff{dropWhile}, \ff{flatMap}, \ff{gather}, \ff{limit}, \ff{peek}, \ff{skip}, \ff{sorted}, \ff{takeWhile}, etc.).
We cannot fuse closures --- lambdas capturing outer variables --- because bytecode places \ff{ILOAD} and \ff{ALOAD} instructions between stream operations, breaking the consecutive pattern our rewriting rules expect.
Similarly, we cannot fuse functional objects (e.g., \ff{new Square()} instead of lambdas) as they compile to \ff{new}/\ff{invokespecial} rather than \ff{invokedynamic}.
Code relying on Reflection to inspect bytecode structure may break, though such practice is uncommon.
Line metadata may be damaged, affecting debugging.

Build time increased by 45 minutes (37\%) for Apache Kafka; merging the current two-component architecture (Java+XSLT and Haskell) into a unified stack may improve this.
We lack micro-level profiling (allocation counts, JIT inlining traces) to clarify the source of performance gains, and we did not compare with JIT-level optimizations such as Graal or HotSpot intrinsics.

Regarding validity: we tested with arrays of \input{_env/numbers.tex} million \ff{int} elements---results may differ with other sizes, types, or objects.
The benchmark by \citet{moller2020eliminating} has low complexity compared to real-world pipelines, lacking method references and autoboxing.
We used a single JVM vendor and hardware configuration with JMH measurements, where JIT behavior variations could affect outcomes.
While Apache Kafka covers nearly one million lines of code, it may not represent all Stream API usage patterns, limiting generalizability to other codebases.
